---
title: "GENE_FAMILY ggtree tree"
subtitle: "Collapsed rectangular phylogram with midpoint rooting"
author: "Your Name"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-overflow: wrap
    code-fold: false
    code-tools: true
    highlight-style: github
    theme: cosmo
    fontsize: 1rem
    linestretch: 1.5
    self-contained: true
execute:
  echo: true
  message: false
  warning: false
  cache: false
---

<!--
TEMPLATE: Collapsed Rectangular Tree (Phylogram and/or Cladogram)
=================================================================
Large trees (100+ tips) with:
  - Pure-clade collapsing by taxonomic group (colored triangles)
  - Branch coloring by taxonomy
  - All visible tips labeled; model species gene names on collapsed triangles
  - No branch capping (branch lengths are never manipulated)

Produces phylogram (branch lengths) and optionally cladogram (topology only).

USAGE: Copy this template into your project's scripts/ directory,
       then adapt the sections marked PROJECT-SPECIFIC.
-->

## Setup

```{r setup}
suppressPackageStartupMessages({
  library(ape)
  library(ggtree)
  library(treeio)
  library(ggplot2)
  library(tidytree)
  library(phytools)
  library(here)
})

# PROJECT-SPECIFIC: Output directory
out_dir <- here("outs", "phylogenetics", "XX_tree_name")
if (dir.exists(out_dir) && length(list.files(out_dir, all.files = TRUE, no.. = TRUE)) > 0) {
  stop("Output destination is not empty; preserve it or use a fresh path")
}
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# PROJECT-SPECIFIC: Declare this QMD plus sourced helpers/configuration.
source_paths <- c("scripts/phylogenetics/XX_tree_name.qmd")
source_files <- file.path(here(), source_paths)
if (!all(file.exists(source_files) & !dir.exists(source_files))) {
  stop("A declared producing file is missing")
}
git_commit <- system2("git", c("-C", shQuote(here()), "rev-parse", "HEAD"), stdout = TRUE)
git_value <- function(arguments) {
  value <- suppressWarnings(system2(
    "git", c("-C", shQuote(here()), arguments), stdout = TRUE, stderr = TRUE
  ))
  if ((!is.null(attr(value, "status")) && attr(value, "status") != 0L) || length(value) != 1) {
    return(NA_character_)
  }
  unname(value)
}
committed_blobs <- vapply(
  source_paths,
  function(relative) git_value(c("rev-parse", shQuote(paste0(git_commit, ":", relative)))),
  character(1)
)
working_blobs <- vapply(
  source_paths, function(relative) git_value(c("hash-object", shQuote(relative))), character(1)
)
if (anyNA(committed_blobs) || anyNA(working_blobs) || !identical(committed_blobs, working_blobs)) {
  stop("Relevant producing files do not match HEAD; make a focused commit before retained execution")
}
cat("Producing commit:", git_commit, "\n")

save_triplet <- function(stem, plot, width, height) {
  ggsave(paste0(stem, ".pdf"), plot, width = width, height = height, units = "in",
         device = grDevices::cairo_pdf)
  ggsave(paste0(stem, ".png"), plot, width = width, height = height, units = "in", dpi = 300)
  ggsave(paste0(stem, ".svg"), plot, width = width, height = height, units = "in",
         device = function(...) svglite::svglite(..., fix_text_size = FALSE))
}
```

## Inputs

```{r inputs}
# PROJECT-SPECIFIC: Tree file and optional metadata
tree_file <- here("data", "phylogenetics", "GENE", "ALIGNMENT.treefile")
# Optional input from a separate lookup producer, outside this tree output directory.
gene_map_file <- here("outs", "gene_lookup", "accession_gene_map.tsv")

tree <- read.tree(tree_file)
cat("Tips:", Ntip(tree), "\n")
if (anyDuplicated(tree$tip.label)) stop("Tree tip labels must be unique; resolve source IDs first")
```

## Taxonomy and label parsing

```{r taxonomy}
# =============================================================================
# PROJECT-SPECIFIC: Tip label parsing
# =============================================================================
# These functions MUST be adapted to match the actual label formats in each tree.
# Inspect tip labels first: head(tree$tip.label, 20)
#
# Common label formats:
#   sp|O95631|NET1_HUMAN           -> species: HUMAN,  accession: O95631 (gene NTN1)
#   tr|Q23158|Q23158_CAEEL         -> species: CAEEL,  accession: Q23158
#   Mus_musculus|10090.Q9R1A3      -> species: Mus_musculus, accession: Q9R1A3
#   Hydra_vulgaris|8692.t25743aep  -> species: Hydra_vulgaris, accession: t25743aep
#
# Key rules:
#   - Model species: resolve actual gene names by accession (gene-lookup skill)
#   - Non-model species: use actual protein/transcript IDs only
#   - NEVER infer gene names from BLAST annotations for non-model species

parse_species_id <- function(label) {
  # UniProt: sp|acc|NAME_SPECIES or tr|acc|NAME_SPECIES
  if (grepl("^(sp|tr)\\|", label)) {
    parts <- strsplit(label, "\\|")[[1]]
    if (length(parts) >= 3) return(sub(".*_", "", parts[3]))
  }
  # Pipe-separated: Species_name|...
  if (grepl("\\|", label)) return(strsplit(label, "\\|")[[1]][1])
  return(label)
}

# Extract the accession/ID, never the UniProt entry mnemonic.
parse_accession <- function(label) {
  if (grepl("^(sp|tr)\\|", label)) return(strsplit(label, "\\|")[[1]][2])
  if (grepl("\\|", label)) {
    after_pipe <- strsplit(label, "\\|")[[1]][2]
    return(sub("^[0-9]+\\.", "", after_pipe))
  }
  return(label)
}

# A single accession may have no known gene; different accessions may share a gene.
load_gene_map <- function(path) {
  if (!file.exists(path)) {
    cat("No accession gene map; retaining accession/ID labels.\n")
    return(character(0))
  }
  gene_map <- read.delim(path, stringsAsFactors = FALSE, colClasses = "character")
  if (!"accession" %in% names(gene_map)) stop("Gene map requires an accession column")
  gene_map$accession <- trimws(gene_map$accession)
  if (anyNA(gene_map$accession) || any(!nzchar(gene_map$accession))) {
    stop("Gene map contains missing/blank accession keys")
  }
  duplicate_keys <- unique(gene_map$accession[duplicated(gene_map$accession)])
  if (length(duplicate_keys) > 0) {
    stop("Duplicate accession keys in gene map; resolve before use: ",
         paste(head(duplicate_keys, 5), collapse = ", "))
  }
  if (!"gene_name" %in% names(gene_map)) gene_map$gene_name <- rep(NA_character_, nrow(gene_map))
  gene_map$gene_name <- trimws(gene_map$gene_name)
  cat("Gene map:", nrow(gene_map), "accessions;",
      sum(!is.na(gene_map$gene_name) & nzchar(gene_map$gene_name)), "with gene names\n")
  setNames(gene_map$gene_name, gene_map$accession)
}

resolve_gene_name <- function(label, species_id = NULL) {
  accession <- parse_accession(label)
  gene <- unname(acc_to_gene[accession])
  if (length(gene) == 1 && !is.na(gene) && nzchar(gene)) return(gene)
  return(accession)
}

report_gene_resolution <- function(labels, mapping) {
  accessions <- vapply(labels, parse_accession, character(1))
  genes <- unname(mapping[accessions])
  mapped <- !is.na(genes) & nzchar(genes)
  cat("Tip gene names:", sum(mapped), "mapped;", sum(!mapped), "accession/ID fallbacks\n")
  if (any(!mapped)) {
    cat("Fallback examples:", paste(head(unique(accessions[!mapped]), 5), collapse = ", "), "\n")
  }
}

acc_to_gene <- load_gene_map(gene_map_file)

# =============================================================================
# PROJECT-SPECIFIC: Taxonomy mapping
# =============================================================================
taxonomy <- list(
  "Demosponges" = c(
    "Amphimedon_queenslandica", "Spongilla_lacustris", "Ephydatia_muelleri",
    "Tethya_wilhelma", "Halichondria_panicea", "Geodia_barretti"
  ),
  "Calcarea/Homoscl." = c(
    "Sycon_ciliatum", "Oscarella_lobularis", "Corticium_candelabrum"
  ),
  "Ctenophora" = c(
    "Mnemiopsis_leidyi", "Pleurobrachia_bachei", "Bolinopsis_microptera"
  ),
  "Cnidaria + Placozoa" = c(
    "Hydra_vulgaris", "Nematostella_vectensis", "Aurelia_aurita",
    "Clytia_hemisphaerica", "Sarsia_tubulosa", "Tripedalia_cystophora",
    "Cladocora_caespitosa", "Trichoplax_sp_H1", "Trichoplax_sp_H2",
    "Trichoplax_sp._H2", "Trichoplax_adhaerens", "Thelohanellus_kitauei",
    "Hydractinia_symbiolongicarpus"
  ),
  "Deuterostomia" = c(
    "Homo_sapiens", "Mus_musculus", "Danio_rerio", "Callorhinchus_milii",
    "Petromyzon_marinus", "Ambystoma_mexicanum", "Xenopus_tropicalis",
    "Ciona_intestinalis", "Strongylocentrotus_purpuratus",
    "Branchiostoma_lanceolatum", "Meara_stichopi", "Waminoa"
  ),
  "Protostomia" = c(
    "Drosophila_melanogaster", "Caenorhabditis_elegans",
    "Parasteatoda_tepidariorum", "Daphnia_pulex", "Hyalella_azteca",
    "Parhyale_hawaiensis", "Platynereis_dumerilii", "Schistosoma_mansoni",
    "Schmidtea_mediterranea", "Prostheceraeus_crozeri",
    "Biomphalaria_glabrata", "Crassostrea_gigas", "Aplysia_californica",
    "Acanthochitona_crinita", "Spadella_cephaloptera"
  )
)

# UniProt suffix -> full species name
uniprot_to_species <- c(
  HUMAN = "Homo_sapiens", MOUSE = "Mus_musculus",
  DROME = "Drosophila_melanogaster", CAEEL = "Caenorhabditis_elegans",
  DANRE = "Danio_rerio", ANOGA = "Anopheles_gambiae",
  RAT = "Rattus_norvegicus", BOVIN = "Bos_taurus",
  CHICK = "Gallus_gallus", XENLA = "Xenopus_laevis",
  SHEEP = "Ovis_aries", ARATH = "Arabidopsis_thaliana",
  DICDI = "Dictyostelium_discoideum"
)

# =============================================================================
# PROJECT-SPECIFIC: Model and focal species
# =============================================================================
# Model species: gene names shown on visible tips AND on collapsed triangle labels
model_species <- c("HUMAN", "MOUSE", "DROME", "CAEEL")
model_species_full <- c("Homo_sapiens", "Mus_musculus",
                        "Drosophila_melanogaster", "Caenorhabditis_elegans")

# Focal non-model species: labeled with protein/gene ID + species abbreviation.
# Protected from collapsing. UPDATE THIS PER TREE.
focal_species_full <- c(
  "Amphimedon_queenslandica", "Spongilla_lacustris", "Ephydatia_muelleri",
  "Tethya_wilhelma", "Halichondria_panicea", "Geodia_barretti",
  "Sycon_ciliatum", "Oscarella_lobularis", "Corticium_candelabrum",
  "Hydra_vulgaris", "Nematostella_vectensis"
)

# =============================================================================
# STYLE PARAMETERS
# =============================================================================
group_colors <- c(
  "Demosponges"         = "#2ca02c",
  "Calcarea/Homoscl."   = "#98df8a",
  "Ctenophora"          = "#9467bd",
  "Cnidaria + Placozoa" = "#ff7f0e",
  "Deuterostomia"       = "#d62728",
  "Protostomia"         = "#1f77b4",
  "Non-metazoan euk."   = "#555555",
  "Unclassified"        = "#8c564b",
  "Mixed"               = "#999999"
)

BRANCH_LINE_WIDTH  <- 0.2
TRIANGLE_ALPHA     <- 0.4
LABEL_SIZE         <- 1.4
LABEL_OFFSET_FRAC  <- 0.005
MARGIN_RIGHT       <- 120       # points — room for labels past plot edge

# Page sizing formula: scales with visible elements after collapsing
# 0.12 in/element keeps labels readable without excess whitespace
INCHES_PER_TIP     <- 0.12
PAGE_WIDTH         <- 8

# Collapsing parameters
COLLAPSE_MIN_PURITY   <- 0.90
COLLAPSE_MIN_SIZE     <- 3
COLLAPSE_MAX_FRACTION <- 0.25

# Species abbreviation: "H. sapiens" from "Homo_sapiens"
abbrev_species <- function(sp_name) {
  parts <- strsplit(sp_name, "_")[[1]]
  if (length(parts) >= 2) return(paste0(substr(parts[1], 1, 1), ". ", parts[2]))
  return(sp_name)
}
```

## Build species lookup and assign tip data

```{r tip-data}
# --- Build species -> group lookup ---
sp_to_group <- list()
for (group in names(taxonomy)) {
  for (sp in taxonomy[[group]]) sp_to_group[[sp]] <- group
}

get_group <- function(species_id) {
  # Try the original species key and its explicit UniProt species mapping.
  species_keys <- c(species_id, unname(uniprot_to_species[species_id]))
  species_keys <- species_keys[!is.na(species_keys) & nzchar(species_keys)]
  for (key in species_keys) {
    if (key %in% names(sp_to_group)) {
      group <- sp_to_group[[key]]
      if (!is.na(group) && nzchar(trimws(group))) return(trimws(group))
    }
  }
  return("Unclassified")
}

report_taxonomy <- function(labels, groups) {
  unclassified <- groups == "Unclassified"
  cat("Taxonomy:", length(groups), "tips;", sum(unclassified), "Unclassified\n")
  print(table(groups))
  if (any(unclassified)) {
    cat("Unclassified examples:", paste(head(labels[unclassified], 5), collapse = ", "), "\n")
  }
}

resolve_species <- function(species_id) {
  if (species_id %in% names(uniprot_to_species)) return(uniprot_to_species[species_id])
  return(species_id)
}

is_model_species <- function(species_id) {
  species_id %in% model_species ||
    resolve_species(species_id) %in% model_species_full
}
```

## Prepare tree

```{r prepare-tree}
# Midpoint root
tree <- phytools::midpoint.root(tree)
cat("Midpoint rooted\n")

# --- Assign data to all tips ---
tip_species_id  <- sapply(tree$tip.label, parse_species_id)
tip_groups      <- sapply(tip_species_id, get_group)
tip_genes       <- mapply(resolve_gene_name, tree$tip.label, tip_species_id)
report_gene_resolution(tree$tip.label, acc_to_gene)
tip_species_full <- sapply(tip_species_id, resolve_species)

tip_data <- data.frame(
  label = tree$tip.label,
  species_id = tip_species_id,
  species_full = tip_species_full,
  group = tip_groups,
  gene = tip_genes,
  stringsAsFactors = FALSE,
  row.names = NULL
)

report_taxonomy(tip_data$label, tip_data$group)

# Display label: "G. species GENE_OR_ID"
tip_data$display_label <- mapply(function(sp_full, gene) {
  sp_abbr <- abbrev_species(sp_full)
  if (!is.na(gene) && nchar(gene) > 0) {
    paste0(sp_abbr, " ", gene)
  } else {
    sp_abbr
  }
}, tip_data$species_full, tip_data$gene)

cat("\nGroup counts:\n")
print(table(tip_data$group))
```

## Tree traversal helpers and node groups

```{r node-groups}
get_tips <- function(tree, node) {
  desc <- getDescendants(tree, node)
  desc[desc <= Ntip(tree)]
}

getDescendants <- function(tree, node) {
  n <- Ntip(tree)
  all_desc <- c()
  queue <- node
  while (length(queue) > 0) {
    current <- queue[1]
    queue <- queue[-1]
    children <- tree$edge[tree$edge[, 1] == current, 2]
    all_desc <- c(all_desc, children)
    queue <- c(queue, children[children > n])
  }
  return(all_desc)
}

# Assign group to every node (tips + internal) for branch coloring
n_tips <- Ntip(tree)
node_group <- rep(NA_character_, n_tips + Nnode(tree))

for (i in seq_len(nrow(tip_data))) {
  tip_idx <- match(tip_data$label[i], tree$tip.label)
  if (!is.na(tip_idx)) node_group[tip_idx] <- tip_data$group[i]
}

# Internal nodes: majority rule (>50% = dominant group, else "Mixed")
for (nd in (n_tips + 1):(n_tips + Nnode(tree))) {
  nd_tips <- get_tips(tree, nd)
  nd_groups <- node_group[nd_tips]
  nd_groups <- nd_groups[!is.na(nd_groups)]
  if (length(nd_groups) > 0) {
    grp_tab <- sort(table(nd_groups), decreasing = TRUE)
    if (grp_tab[1] / sum(grp_tab) > 0.5) {
      node_group[nd] <- names(grp_tab)[1]
    } else {
      node_group[nd] <- "Mixed"
    }
  }
}
```

## Collapse pure taxonomic group clades

Collapse monophyletic clades where 90%+ of tips belong to the same taxonomic
group. The `collapse_groups` parameter controls which groups are eligible for
collapsing — by default only Bilateria are collapsed.

```{r collapse}
find_collapsible_by_group <- function(tree, tip_data,
                                       min_purity = COLLAPSE_MIN_PURITY,
                                       min_size = COLLAPSE_MIN_SIZE,
                                       max_fraction = COLLAPSE_MAX_FRACTION,
                                       collapse_groups = NULL) {
  n_tips <- Ntip(tree)
  max_size <- floor(n_tips * max_fraction)
  node_ids <- (n_tips + 1):(n_tips + Nnode(tree))

  clade_info <- lapply(node_ids, function(nd) {
    nd_tips <- get_tips(tree, nd)
    nd_groups <- tip_data$group[nd_tips]
    grp_tab <- sort(table(nd_groups), decreasing = TRUE)
    list(node = nd, size = length(nd_tips),
         dominant_group = names(grp_tab)[1],
         purity = grp_tab[1] / length(nd_tips),
         tip_ids = nd_tips,
         group = names(grp_tab)[1])
  })

  candidates <- Filter(function(x) {
    group_ok <- if (is.null(collapse_groups)) TRUE else
                x$dominant_group %in% collapse_groups
    x$purity >= min_purity & x$size >= min_size & x$size <= max_size & group_ok
  }, clade_info)

  # Sort by size descending (collapse biggest pure clades first)
  candidates <- candidates[order(sapply(candidates, `[[`, "size"),
                                  decreasing = TRUE)]

  # Greedy non-overlapping selection
  selected <- list()
  collapsed_tips <- integer(0)
  for (cand in candidates) {
    if (any(cand$tip_ids %in% collapsed_tips)) next
    selected <- c(selected, list(cand))
    collapsed_tips <- c(collapsed_tips, cand$tip_ids)
  }
  return(selected)
}

# PROJECT-SPECIFIC: Which groups to collapse (NULL = all groups eligible)
clades_to_collapse <- find_collapsible_by_group(tree, tip_data,
                                                 collapse_groups = c("Bilateria"))

cat("Collapsing", length(clades_to_collapse), "clades:\n")
for (cl in clades_to_collapse) {
  cat(sprintf("  Node %d: %d tips (%s, %.0f%% pure)\n",
              cl$node, cl$size, cl$dominant_group, cl$purity * 100))
}

collapsed_tip_ids <- unlist(lapply(clades_to_collapse, function(cl) cl$tip_ids))
visible_tip_ids <- setdiff(seq_len(Ntip(tree)), collapsed_tip_ids)
n_visible <- length(visible_tip_ids) + length(clades_to_collapse)

# --- Page height from visible element count ---
PAGE_HEIGHT <- max(8, n_visible * INCHES_PER_TIP)
cat(sprintf("\nVisible elements: %d tips + %d collapsed = %d total\n",
            length(visible_tip_ids), length(clades_to_collapse), n_visible))
cat(sprintf("Page: %.1f x %.1f in (%.2f in/element)\n",
            PAGE_WIDTH, PAGE_HEIGHT, INCHES_PER_TIP))
```

## Build collapse labels with model species gene names

Collapsed triangle labels include the group name, tip count, and gene names
of any model species (human, mouse, fly, worm) inside the collapsed clade.
This ensures key gene family members remain visible even when collapsed.

```{r collapse-labels}
# Filter out UniProt accession-like strings — keep only real gene symbols.
# tr| entries produce accession strings (e.g., A0A061AD21) instead of gene names.
is_gene_symbol <- function(name) {
  if (is.na(name) || nchar(name) == 0) return(FALSE)
  if (grepl("^A0A", name)) return(FALSE)
  if (grepl("^[OPQ][0-9][A-Z0-9]{3}[0-9]$", name)) return(FALSE)
  if (grepl("^[A-NR-Z][0-9][A-Z][A-Z0-9]{2}[0-9]$", name)) return(FALSE)
  if (grepl("^[A-Z][0-9][A-Z0-9]{4,}", name)) return(FALSE)
  return(TRUE)
}

# Extract model species gene SYMBOLS (not accessions) from a collapsed clade
get_model_genes_in_clade <- function(cl, tip_data) {
  clade_tips <- tip_data[cl$tip_ids, ]
  model_rows <- clade_tips[
    clade_tips$species_id %in% model_species |
    clade_tips$species_full %in% model_species_full, ]
  if (nrow(model_rows) == 0) return(NULL)
  genes <- model_rows$gene[!is.na(model_rows$gene) & nchar(model_rows$gene) > 0]
  genes <- genes[sapply(genes, is_gene_symbol)]
  if (length(genes) == 0) return(NULL)
  return(unique(sort(genes)))
}

# Build label: "Group (N tips: GENE1, GENE2, ...)" or "Group (N tips)" if none
format_collapse_label <- function(cl, tip_data) {
  genes <- get_model_genes_in_clade(cl, tip_data)
  if (is.null(genes)) {
    paste0(cl$dominant_group, " (", cl$size, " tips)")
  } else {
    gene_str <- paste(genes, collapse = ", ")
    paste0(cl$dominant_group, " (", cl$size, " tips: ", gene_str, ")")
  }
}

# Preview what labels will look like
for (cl in clades_to_collapse) {
  cat(format_collapse_label(cl, tip_data), "\n")
}
```

## Render phylogram

```{r phylogram}
#| fig-width: 8
#| fig-height: 20

p <- ggtree(tree, layout = "rectangular",
            size = BRANCH_LINE_WIDTH, aes(color = group)) %<+% tip_data

# Inject node group assignments for branch coloring
p$data$group <- node_group[p$data$node]
p$data$group[is.na(p$data$group)] <- "Mixed"

# --- CRITICAL: Pre-compute positions BEFORE collapse ---
# collapse() modifies p$data coordinates, so we must capture them first.
# Use match() on node column — do NOT assume p$data rows are ordered by node ID.
pre_data <- p$data
x_range <- diff(range(pre_data$x, na.rm = TRUE))
label_offset <- x_range * LABEL_OFFSET_FRAC

# Tip labels for non-collapsed tips
visible_rows <- tip_data[visible_tip_ids, ]
tip_node_ids <- match(visible_rows$label, tree$tip.label)
data_row_idx <- match(tip_node_ids, pre_data$node)
tip_label_df <- data.frame(
  x = pre_data$x[data_row_idx],
  y = pre_data$y[data_row_idx],
  tip_label = visible_rows$display_label,
  group = visible_rows$group,
  stringsAsFactors = FALSE
)

# Collapse triangle labels — position at max(x) of descendant tips (triangle
# tip), NOT at the internal node x (triangle base). This is a critical fix:
# the internal node sits at the base of the triangle, but the label should
# appear at the point where the triangle extends to.
collapse_label_df <- do.call(rbind, lapply(clades_to_collapse, function(cl) {
  nd_row <- match(cl$node, pre_data$node)
  tip_x_max <- max(pre_data$x[cl$tip_ids])
  tip_y <- pre_data$y[nd_row]
  lbl <- format_collapse_label(cl, tip_data)
  data.frame(x = tip_x_max, y = tip_y, tip_label = lbl,
             group = cl$group, stringsAsFactors = FALSE)
}))

# --- Apply collapse (colored triangles) ---
for (cl in clades_to_collapse) {
  color <- group_colors[cl$group]
  if (is.na(color)) color <- "#999999"
  p <- p |> collapse(node = cl$node, mode = "max",
                     fill = color, alpha = TRIANGLE_ALPHA, color = color,
                     linewidth = BRANCH_LINE_WIDTH)
}

# --- Add labels ---
p <- p +
  geom_text(data = tip_label_df,
            aes(x = x + label_offset, y = y, label = tip_label, color = group),
            hjust = 0, size = LABEL_SIZE,
            show.legend = FALSE, inherit.aes = FALSE) +
  geom_text(data = collapse_label_df,
            aes(x = x + label_offset, y = y, label = tip_label, color = group),
            hjust = 0, size = LABEL_SIZE * 0.85, fontface = "italic",
            show.legend = FALSE, inherit.aes = FALSE) +
  scale_color_manual(
    values = group_colors, na.value = "#999999", name = "Taxonomic Group"
  ) +
  scale_y_continuous(expand = c(0.005, 0.005)) +
  coord_cartesian(clip = "off") +
  theme_tree() +
  theme(
    plot.margin = margin(5, MARGIN_RIGHT, 5, 5),
    legend.position = "bottom",
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 10, face = "bold")
  ) +
  guides(color = guide_legend(override.aes = list(linewidth = 2)))

save_triplet(file.path(out_dir, "GENE_ggtree_rect"), p, PAGE_WIDTH, PAGE_HEIGHT)

p
```

## Session info

```{r build-info}
expected_outputs <- paste0(file.path(out_dir, "GENE_ggtree_rect"), c(".pdf", ".png", ".svg"))
if (any(!file.exists(expected_outputs))) stop("Expected tree output is missing")
final_working_blobs <- vapply(
  source_paths, function(relative) git_value(c("hash-object", shQuote(relative))), character(1)
)
if (anyNA(final_working_blobs) || !identical(committed_blobs, final_working_blobs)) {
  stop("Relevant producing files changed during execution")
}
sessionInfo()
writeLines(
  c(
    "completion: complete",
    "script: scripts/phylogenetics/XX_tree_name.qmd",
    paste("date:", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
    paste("commit:", git_commit),
    paste("environment:", R.version.string),
    "invocation: quarto render scripts/phylogenetics/XX_tree_name.qmd --output-dir outs/phylogenetics/XX_tree_name/",
    paste("input:", tree_file, paste0("| md5=", unname(tools::md5sum(tree_file)))),
    paste("source:", source_paths, "| content verified against commit"),
    paste("output:", expected_outputs)
  ),
  file.path(out_dir, "BUILD_INFO.txt")
)
```
